### 2012-07-17 paper_draft.tex
Ironed out sentences, shortened a lot, re-arranged some stuff

If the genotypes for more than one individual are known, the analysis of
allele frequencies can be used to associate SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
(GWAS) make use of statistics to compare the alleles of patients to the alleles of healthy controls. This
enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising. Over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
get an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
alone had over 100,000 customers \cite{23andMe2011}
- the company recognizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
traits and diseases. With the consent of the customer the data is used for association studies. 23andMe has published several 
articles in which it replicates known findings, but also finds new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}. 
Over 30,000 23andme-customers have participated in those association studies.

Although companies like 23andMe are willing to use the data for scientific research,
it is not easy for individual scientists to access the data. 
The generation of biomedical data by private companies raises concerns about privacy \cite{23andMe2012}, 
liabilty and consent \cite{Caulfield2011}. 
Nevertheless, individual customers are willingly sharing their data. Most do so by uploading their data to
their personal website or to open software repositories like \textit{GitHub}. 
This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all 
the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is 
most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
like the \textit{Personal Genome Project}, still do not allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.

We designed an online survey to evaluate the support in the personal genetics community for a crowdsourced online platform. Here we present the results of the survey in combination with the openSNP project. 
The openSNP project is an online platform, which makes it possible for DTC customers to share genotypic and phenoytypic information and connect it with external data sources. 
The data is made available to research via a Creative-Commons-Zero licence.

@@

The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers. 
DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable GWAS for researchers,
but the data can also be used for other research and in any other context.  
Users can create a personal profile, discuss SNPs and phenotypes on the 
platform using a simple commenting system, or send each other private messages.

\subsubsection*{Sharing genotypic information}
Currently users can upload their genotypings from the companies 23andme, deCODEme and FamilyTreeDNA via a webinterface to the openSNP 
project. The uploaded data is published under the Creative Commons Zero-license, 
which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - 
allows a complete reuse of the data without any constraints.
Between the start of openSNP on 09/27/2011 and 12/18/2012, 214 people have signed 
up with openSNP, and 79 genotyping files were made available. The openSNP 
database lists 69,486,471 SNPs, which are distributed over 1,938,603 unique SNP-IDs.
Figure \ref{Figure1_label} depicts the increase of users and genotyping files over time.

\subsubsection*{Crowdsourcing phenotypes}
Beside uploading genotypic data, users may create new phenotypes that are not yet 
listed by openSNP. 
The specification of these phenotypes is open and not limited 
to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
the expression and spelling of the same phenotype or variation. We implemented an 
autocompletion-feature, which helps users to reuse already entered phenotypes.
Users are encouraged to list as many phenotypes as possible through a simple 
achievement system, rewarding users that upload their data and enter phenotypic 
information with small badges that are shown on their profile pages.

In the same timeframe that was mentioned above, all users combined have 
entered a total of 675 variations on 47 different phenotypes with those variations being 
the different values on a given trait or phenotype. See figure \ref{Figure1_label} for the increase of phenotypic information over time.

@@

\subsubsection*{Connection to external services}
In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. 
Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge by every user (Figure \ref{Figure3_label}). For usability reasons, 
SNPs are ranked by the amount of information gathered by the external services. The external services themselves are ranked by how easily users can access information 
from these sources. The SNPedia entries are given the highest impact, as those are already manually curated, followed by open access publications out of 
the Public Library of Science. Lowest values are given to the Mendeley results, as those are not necessary freely available to every user. 
An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.

@@

\subsection*{Survey issues}
As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, 
the results do not reflect the general population, but overrepresent those people most likely to be interested in
a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology. 
As we wanted to find out about interest in crowdsourcing GWAS and about how people who had purchased DTC genetic 
testing felt about sharing their data with third parties, this approach was warranted.

\subsection*{GWAS and open data}
Although prices of exome or even full genome sequencing are dropping rapidly, GWAS tend to stay considerably cheaper 
and offer a possibility of getting insight into genetic variations, for example SNPs which are linked to different traits 
on a population level.

Despite this benefit in terms of cost GWAS can only detect correlations of SNPs with those traits and do not allow 
inference on the cause for any correlation. Furthermore, GWAS need a large enough sample size, 
which has to be statistically analysed by using sound methods. 
Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.

@@

By crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other 
open platform on the Internet, namely the need to trust users regarding the data they upload and enter on openSNP. 
Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation, 
with users entering data in different measurement systems. While we try to suggest similar entries to the users, 
there are some cases where users will not follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise.

There are two possible solutions to this problem: one could only allow some (trusted) users to enter new phenotypes or one 
could make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
variations that have not been available from the start. Both methods have a disadvantage: In either case it makes it harder 
for users to enter their data and thereby raises the bar for participation, which ultimately could lead to less data entered. 
Facing this trade-off, we decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS 
with the data to perform more quality control.

Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot 
rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing, from which an even smaller subset 
is willing to publish the results, along with phenotypic information of medical relevance.

With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic 
data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. 
Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
While we do not have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising.

\subsection*{Privacy, health implications and ethical considerations}
As we have found in the survey on sharing such results, many of these customers are also willing to share their results with scientists and 
the public to help scientific progress, although people who have taken DTC genetic testing are considering the privacy implications that come 
with openly sharing those results.

%The advent of DTC genetic testing has led to new ethical and social issues. 
Much of the critique on DTC genetic testing focusses on the practice 
of delivering medical information without consulting a physician or genetic counselor to help patients/customers make sense of the information 
and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}. Less attention has been paid to the privacy implications 
that come with this cheap way of obtaining genetical information \cite{Caulfield2011,Joh2011}.

With a move towards making one's own genetic and medical information public, it becomes clear that the issue of privacy needs to get a closer look. 
Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments 
or advertisers might misuse the information. Policy makers start to react to those changes by introducing laws like the 
\textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.

We openly address the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during 
the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
that can arise by publishing their data. To further improve this process we are working on implementing informed consent-processes,
like the one by \textit{Consent for Research} (http://www.weconsent.us).

@@

The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
(planned) customership of a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
but only if they did not share my medical information with anybody else" and No".

The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
information with third parties. The scale went from 1 = strongly disagree over 3 = neutral to  5 = strongly agree. Motivations queried for sharing data 
were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
"because of the breach of my privacy" and " because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.

@@

A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure \ref{Figure4_label}. The source code of openSNP is 
published under Creative Commons BY-SA 3.0 and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

>>>>

If the genotypes for more than one individual are known, the analysis of
allele frequencies can be used to associate SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
(GWAS) make use of statistics to compare the alleles of patients to the alleles of healthy controls. This
enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising. To date, over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits \cite{Hindorff2009}. 
%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

GWAS are not limited to the scientific community. Since 2006, companies like 23andMe, deCODEme or FamilyTreeDNA have been offering Direct-To-Consumer (DTC) genetic testing. 
These companies use DNA microarrays to screen for around 0.5 to 1 million SNPs spread over the human genome. In return, customers 
receive an analysis of the results, as well as a file that includes the customer's raw individual genotypes. In 2011, 23andMe 
alone had over 100,000 customers \cite{23andMe2011}
- the company realizes the potential to perform GWAS with this amount of data by using surveys to ask their customers about 
traits and diseases. With the consent of the customer the data is used for association studies. 23andMe has published several 
articles in which it replicates known findings, but also finds new associations for Parkinson's Disease \cite{Eriksson2010, Do2011}. 
So far, over 30,000 23andme-customers have participated in those association studies, which proves that this data-source has potential for other researchers.

The generation of biomedical data by private companies raises concerns about privacy \cite{23andMe2012}, 
liability and consent \cite{Caulfield2011}. 
Nevertheless, in some instances individual customers are willingly sharing their data. Most do so by uploading their data to
their personal website or to open software repositories like \textit{GitHub}. 
This data is scattered and unorganized, making it hard to use. While projects like SNPedia try to keep track of all 
the publicly available genotyping files \cite{Cariaso2011}, they usually do not provide the information necessary to perform GWAS, as the phenotypic information is 
often not attached to the genetic information. Projects that attach the phenotype to the genetic information, 
like the \textit{Personal Genome Project}, still do not allow for an easy re-use of the data, as they lack an advanced programming interface (API) 
or other methods by which researchers could download the data. Additionally, not every customer of DTC genetic testing can participate in the \textit{Personal Genome Project}.

Here, we present the results of a survey designed to evaluate the support in the personal genetics community for a crowdsourced online platform. 
We also present openSNP, an online platform which enables DTC customers to share genotypic and phenoytypic information, as well as receive additional information on their genotypes. 
The genotypes are made available to researchers via the open Creative-Commons-Zero licence.

@@

The openSNP project (http://opensnp.org) is an open, crowdsourced online platform for both DTC customers and researchers interested in performing GWAS. 
DTC customers are encouraged to share their genotyping results and connect them with phenotypes to enable access for researchers.
Users can create a personal profile, discuss SNPs and phenotypes on the 
platform using a simple commenting system, or send each other private messages.

\subsubsection*{Sharing genotypic information}
Currently users can upload their genotypings from the companies 23andme, deCODEme and FamilyTreeDNA via a webinterface to the openSNP 
project. There is experimental support for openSNP-exomes. The uploaded data is published under the Creative Commons Zero-license, 
which - in accordance with the Panton Principles (doi:10.1371/journal.pbio.1001195) - 
allows a complete reuse of the data without any constraints.
Between the start of openSNP on 09/27/2011 and 12/18/2012, 214 people have signed 
up with openSNP, and 79 genotyping files were made available. The openSNP 
database lists 69,486,471 SNPs of which 1,938,603 are unique SNPs.
Figure \ref{Figure1_label} depicts the increase of users and genotyping files over time.

\subsubsection*{Crowdsourcing phenotypes}
Users are able to create new phenotypes that are not yet 
listed by openSNP. 
The specification of these phenotypes is open and not limited 
to pre-defined categories. To reduce the amount of manual data curation, openSNP tries to harmonize 
the expression and spelling of the same phenotype or variation. We implemented an 
autocompletion-feature, which helps users to reuse already entered phenotypes.
Users are encouraged to list as many phenotypes as possible through a simple 
achievement system, rewarding users that upload their data and enter phenotypic 
information with small badges that are shown on their profile pages.

In the same timeframe as above, all users combined have 
entered a total of 675 variations on 47 different phenotypes with those variations being 
the different values on a given trait or phenotype. See figure \ref{Figure1_label} for the increase of phenotypic information over time.

@@

\subsubsection*{Connection to external services}
In order to provide users with relevant information on their respective genotypes, openSNP scans databases of the scientific literature for specific SNPs. 
A total number of 15,229 documents relevant to the SNP-IDs listed in openSNP could be found in the databases of Mendeley, the Public Library of Science and SNPedia. 
Of the primary literature, 25 \% are released in open access journals and can be accessed free of charge (Figure \ref{Figure3_label}). For usability reasons, 
SNPs are ranked by the amount of information gathered by the external services. The external services themselves are ranked by how easily non-scientists can understand information 
from these sources. The SNPedia entries are given the highest impact, as those are already manually curated and summarized in plain English, followed by open access publications out of 
the Public Library of Science. Lowest values are given to the Mendeley results, as those are not necessarily freely available to every user. 
An entry on SNPedia is valued 2.5 times as high as a PLoS publication and 5 times as high as a Mendeley entry.

@@

\subsection*{Survey issues}
As the survey was taken online by voluntary participants and was mainly spread in the personal genetics community, 
the results do not reflect the general population, but over-represent those people most likely to be interested in
a project such as openSNP: customers of DTC genetic testing companies and people with a high interest in biology.

\subsection*{Privacy, health implications and ethical considerations}

%The advent of DTC genetic testing has led to new ethical and social issues. 
Much of the critique on DTC genetic testing focusses on the practice 
of delivering medical information without consulting a physician or genetic counsellor to help patients/customers make sense of the information 
and to put the new knowledge to good use \cite{Hauskeller2011,Hogarth2008,Wasson2009}.

As we have found in the survey on sharing such results, many DTC customers are willing to share their results with the public to help scientific progress, without forgetting about the privacy implications that come with openly sharing genetic information. There is a variety of ethical and privacy implications when it comes to DTC genetic testing\cite{Caulfield2011,Joh2011}.

Our survey has shown that people are concerned about their privacy and fear that stakeholders like employers, insurance companies, governments 
or advertisers might misuse the information. Policy makers start to react to those changes by introducing laws like the 
\textit{Genetic Information Non-Discrimination Act} in the United States or the \emph{Gendiagnostikgesetz} in Germany to minimize the impact of
widely available genetic information. DTC genetic testing companies themselves also try to educate their customers about the risks of releasing genetic data.

OpenSNP openly addresses the problem of privacy implications that come with releasing genetic data twice, during registration for openSNP and during 
the upload of the DTC genetic testing results. Users have to confirm that they have read and understood the disclaimer about possible side-effects 
that can arise by publishing their data. To further improve this process we are working on implementing an informed consent-processes,
similar to \textit{Consent for Research} (http://www.weconsent.us).

\subsection*{GWAS and open data}
Although prices of exome or even full genome sequencing are dropping rapidly, GWAS are still considerably cheaper. However, GWAS can only detect correlations of SNPs with those traits and do not allow 
inference on the cause for any correlation. Furthermore, for a statistically sound analysis GWAS need a large enough sample size. Nevertheless, GWAS are still frequently used and new associations are found \cite{10.1371.journal.pone.0031470,10.1371.journal.pone.0030309,10.1371.journal.pone.0029848}.

@@

By crowdsourcing the acquisition of genetic and phenotypic data, openSNP faces the same problems as any other 
open platform on the Internet, namely the need to trust users regarding the data they upload and enter on openSNP. 
Additionally, the quality of the data varies, especially in terms of accuracy on the phenotypic variation, 
with users entering data in different measurement systems. Another problem with user-entered data is the frequent switching between categorical and continuous phenotypes of users - for example, some users entered the specific value of their height, while other users entered their height according to a category like "150cm to 160cm".

While we try to suggest similar entries to the users, 
there are some cases where users will not follow those suggestions, so duplicates or similar phenotypes or variations in traits may arise. There are two possible solutions to this problem: The first one would be to only allow a trusted subset of users to enter new phenotypes. The other one 
would be to make users enter all possible variations of a phenotype while creating a new phenotype, so that later users cannot add 
variations that have not been available from the start.

In both cases it makes it harder 
for users to enter their data which raises the bar for participation.
We decided to keep data entry as easy as possible, at the cost of forcing users who want to perform GWAS with the data to perform more quality control.

Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP: only a subset of people buy DTC genetic testing, from which an even smaller subset is willing to publish the results, which can potentially lead to skewed GWAS-results.

With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic 
data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to use in GWAS. 
Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. 
While there is not enough data uploaded to perform a statistically sound GWAS yet, this will be possible in the future, as user numbers continue to rise.

@@

The survey included demographics such as age, chromosomal sex and ethnicity of the participants. Furthermore, it included questions on their 
(planned) customer-ship of a DTC company. If the participants already were customers, they were also asked if they were already sharing their genetic and phenotypic data. 
All participants were asked if they would share their genetical or phenotypic information with their DTC company, possible answers were "Yes", "Yes, 
but only if they did not share my medical information with anybody else" and No".

The survey also asked some scaled questions which measured how strongly participants agreed/disagreed with different reasons for sharing or not sharing their 
information. The scale went from 1 = strongly disagree to  5 = strongly agree. Motivations queried for sharing data 
were "because I want to help scientists with their research", "because of possible personal benefits (e.g. getting treatments for a disease I have, 
possibility of new medication, etc.)", "because it may deliver advertising that is relevant to me" and "out of curiosity". Motivations queried for not sharing 
data were "because advertisers could use the information for targeted campaigns", "because of possible negative consequences for closely related persons", 
"because of the breach of my privacy" and " because of the fear of discrimination (e.g. by the employer, the state, some insurance company)". 
Additionally, participants had the possibility of giving their own reasons for sharing or not sharing their data.

@@

A flowchart of all services incorporated in openSNP and of all the ways users can upload or access the data is given in figure \ref{Figure4_label}. The source code of openSNP is 
published under the MIT-license and can be downloaded at http://github.com/gedankenstuecke/snpr. The genetical and phenotypical data is licensed under Creative Commons Zero. 
% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}
